### 2019-05-16 2-Chapters/2-Chapter/chapter2.tex
Just hard :cry:

\paragraph{Abstract.}
%
In this chapter, we present the common base of all the models studied in this thesis:
the single-player and stationary stochastic MAB model.
We focus on decision-making models with a finite number of resources, called arms, and on stochastic models, where an arm is associated with a one-dimensional distribution.
In all this thesis, we consider one-dimensional exponential families, and in particular Bernoulli distributions.
%
We review quickly the possible applications of bandits, that range from recommender systems and information retrieval to healthcare, finance. In the rest of this thesis, we focus on cognitive radio, an important field of research where bandits can be applied with success.

We then explain the measure of performance used in this thesis, defined as the regret and written $R_T^{\cA}$ for algorithm $\cA$ and horizon $T$. The regret compares the performance of the decision-making player and the oracle who knows in advance the best arm and always selects it.
Other measures of performance, such as best-arm identification, are not used later on in this document, but they are also shortly discussed.
%
We review the main theoretical results from famous previous works, that give a lower-bound on the regret of any algorithm, essentially showing that we cannot find an algorithm achieving a sub-logarithmic regret, in this stochastic MAB model.

We give an overview of the main families of algorithms designed to tackle the stochastic MAB model, and we mainly focus on the \UCB, \klUCB{} and Thompson sampling algorithms, as they are the two we used in the next chapters.
% Part of the work presented in Chapter~\ref{chapter:4} also used the Thompson sampling algorithm, but no theoretical analyses were done about it.
When presenting our library SMPyBandits in Chapter~\ref{chapter:3}, designed to perform numerical simulations of MAB problem, we use it to compare the empirical performances of the most famous and most efficient algorithms, on different problems.
%
In the point-of-view of numerical simulations or real-world applications of bandits on embedded systems, one must also take care of two other measures of performance: time and memory complexities, which can vary greatly between different families of algorithms.
We also give an empirical comparison of different algorithms for these two measures in Chapter~\ref{chapter:3}.

We also gives regret upper-bounds of the two main algorithms used in this thesis, \UCB{} and \klUCB, to highlight the difference between their finite-time regret upper-bounds. They both attain a logarithmic regret, proving their order optimality (\ie, their regret upper-bound asymptotically matches the lower-bound), moreover \klUCB{} is known to be optimal for a wide range of arms distributions.
% A sketch of the proof of both results is given in Appendix.

@@

\paragraph{Main references for the curious reader.}
%
For more details and a more pedagogical introduction to multi-armed bandits, we suggest to the interested reader to read the short text-book \cite{Slivkins2019}.
Another excellent but reasonably short survey is \cite{Bubeck12}, while the more recent book \cite{LattimoreBanditAlgorithmsBook} is the most complete resource about bandit algorithms.
Finally, we recommend \cite{bouneffouf2019survey} for a short but good survey on applications of MAB.

@@

Multi-Armed Bandits models were first introduced by Thompson as early as 1933 in \cite{Thompson33}, and later studied from the 1950s by Robbins \cite{Robbins52}.
This family of models was first proposed for clinical trials, and later applied to a wide range of different problems.
The name \emph{multi-armed bandits} comes from the one-armed bandit found in casinos, see Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem__step0} below.
% The seminal paper by Lai and Robbins \cite{LaiRobbins85} was the first to introduce the measure of performance widely used in the literature, the notion of \emph{regret}, and to analyze a lower-bound on the performance of any algorithm against a fixed problem.
% Since then, many efficient algorithms were proposed, XXX

%
A MAB is a model of a decision-making game, where a \emph{player} has to sequentially select one action in a (usually finite) set of actions, and only receives a (random) feedback about the action she selected, also called \emph{reward}.
More formally, we consider $K\in\N, K\geq2$ arms, and the player's decision at time $t$ is denoted $A(t)\in\{1,\dots,K\}$, after what she receives a reward $r_{A(t)}(t)\in\R$, and so on for $t\in\N^*$.
The goal of the player is to maximize its sum of received rewards, from time $t=1$ to $t=T$ if the game is played at an horizon $T$. The difficulty comes from balancing the trade-off between \emph{exploration}, because she has to observe as much as possible all the arms to get more knowledge about the distributions of their rewards, and \emph{exploitation}, because she also has to select as much as possible the best arm.
The MAB model is a famous example of a reinforcement learning model, where the decision-making process has to adapt to an unknown environment using noisy observations and a discrete time, alternating between decisions and observations.
We illustrate this cycle in Figure~\ref{fig:2:ReinforcementLearningCycleMABmodel} below.

\tikzstyle{block} = [align=center, draw, fill=gray!25, rectangle, minimum height=3em, minimum width=6em]
\begin{figure}[h!]
    \centering
    \resizebox{0.40\textwidth}{!}{
        \begin{tikzpicture}[auto,node distance=5cm,>=latex,scale=2]
            %
            % We start by placing the blocks
            \node [block] (player) at (0,0) {Player};
            % We draw an edge between the player and system block to
            \node [block] (environment) at (2,0) {MAB problem};
            % Once the nodes are placed, connecting them is easy.
            \draw [->] (player) to[bend left=90] node[pos=0.5] {Discrete action $k=A(t) \in\{1,\dots,K\}$} (environment);
            \draw [->] (environment) to[bend left=90] node[pos=0.5] {Real-valued reward $r_k(t) \in \R$} (player);
            %
        \end{tikzpicture}
    }
\caption{Reinforcement learning cycle in a MAB model, for time steps $t=1,\dots,T$.}
\label{fig:2:ReinforcementLearningCycleMABmodel}
\end{figure}

The first application of MAB models was for clinical trials, were an arm represent a treatment, and the distribution associated with such treatment can be a Bernoulli distribution: a reward of $0$ means the drug did not heal the disease, and a reward of $1$ indicates a success. The mean of an arm, in this application, represents the mean success rate of a treatment, and the goal of a doctor in a clinical trial is to identify the best treatment, \ie, the arm with highest mean, in a number of trials as short as possible.
%
The focus of this work is on cognitive radio and IoT networks, where arms can represent wireless orthogonal channels, but more generally any resource characterizing the communication of a device to a gateway (\eg, spreading factor for LoRa, power allocation for NOMA etc).
%
After introducing formally the notations used in all this thesis, we review below possible applications of multi-armed bandits.

@@

In all this thesis, we focus on the model with finite arms, and binary or real-valued stochastic rewards, that is $A(t)\in\{1,\dots,K\}$ for a fixed and known number of arms $K\in\N,K\geq2$, and $r_{A(t)}(t)\in\R$.
%
Restricting to models with a finite number of arms
rules out an entire domain of the recent research on MAB, in particular we do not consider linear nor contextual bandits (for more details, see Parts V and VI of \cite{LattimoreBanditAlgorithmsBook}).

We also restrict to stochastic rewards, that is we associate a real-valued distribution to each arm, denoted $\nu_k$ for arm $k\in\{1,\dots,\}=[K]$.
Rewards are stochastic and stationary, meaning that $(r_k(t))_{t\in\N^*}$ is independent and identically distributed (\iid), $r_k(t) \sim \nu_k$ for any $t\geq1$.

@@

The webpage looks like Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
The arms follow Bernoulli distributions, \ie, $\nu_k = \mathcal{B}(\mu_k)$, of unknown means $\mu_k\in[0,1]$, and your goal in this interactive demonstration is to obtain the highest possible cumulated reward in $100$ steps, $\sum_{t=1}^{100} r_{A(t)(t)$.
Your decisions have to be made sequentially: at time $t$, you pick one of the arms, $A(t) \in\{1,2,3,4,5\}$, then the demo shows the random reward obtained from this (virtual) casino machine (in \textcolor{gold}{yellow}), \ie, the binary value $r_{A(t)}(t)\in\{0,1}$, that is sampled \iid{} from $\nu_k$.
%
The UI of the demo also shows the current value of $t$ (``total plays'') and $\sum_{s=1}^t r_{A(s)}(s)$ the ``total reward''.
For each arm, we show the sum of rewards obtained from that arm, \ie, $X_k(t) = \sum_{s=1}^t r_k(s) \mathbbm{1}(A(s) = k)$, in the ``Rewards'' line, and the number of pulls of that arm, \ie, $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$ in the ``Pulls'' line.
%
The demo also shows the estimated probability of each arm, that is $\widehat{\mu_k}(t) = X_k(t) / N_k(t)$ (when $N_k(t)>0$), in the ``Estimated probs'' line.

We first illustrate in Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} the current state of the game at time $t=24$.
% Currently at time $t=24$ (out of $T=100$ total time steps),
At this step, the player has collected a sum of rewards of $14$, by observing $X(t) = [6,2,2,2,2]$ rewards of value $1$ in the $K=5$ different arms. Arms were sampled $N(t) = [8,4,4,4,4]$ times, meaning that the value $0$ was seen respectively $[2,2,2,2,2]$ times, and currently arm $1$ appears to be the best one. The true means of the arms are $\bm{\mu}=[0.6, 0.2, 0.55, 0.7, 0.5]$, and (much) more samples are needed before the player can accurately identify arm $4$ as the best arm.
%
We also illustrate in Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem__step100} the result of an example of run, when the player was following the \UCB{} algorithm from \cite{Auer02}.
After $T=100$ steps, the player obtained a cumulated reward of $56$, by playing mostly arms $4,3,5,1,0$ (in decreasing order of number of plays). The empirical means $\widehat{\mu_k}(T)$ correctly identify the best arm (arm $4$), but do not correctly rank the arms as arms $1$ and $3$ obtained means of $\widehat{\mu_1}(T) = 0.5 < \widehat{\mu_3}(T)=0.6$ but the true means are $\mu_3 = 0.55 < \mu_1 = 0.6$.

@@

\begin{figure}[h!]  % [htbp]
    \centering
    \includegraphics[width=0.95\linewidth]{2-Chapters/2-Chapter/Images/example_of_a_5_arm_bandit_problem.png}
    \caption{Screenshot of the demonstration available online on my website, for a current step of $t=24$. See \href{https://perso.crans.org/besson/phd/MAB\_interactive\_demo/}{\texttt{perso.crans.org/besson/phd/MAB\_interactive\_demo/}}.}
    \label{fig:2:example_of_a_5_arm_bandit_problem}
\end{figure}

\begin{figure}[h!]  % [htbp]
    \centering
    \includegraphics[width=0.95\linewidth]{2-Chapters/2-Chapter/Images/example_of_a_5_arm_bandit_problem__step100.png}
    \caption{Screenshot of the demonstration, after the end of the game, after $t=T=100$ steps.}
    \label{fig:2:example_of_a_5_arm_bandit_problem__step100}
\end{figure}

@@

% Just bounded, in $[a,b]$ but without loss of generality in $[0,1]$ etc.
\paragraph{The normalization trick.}
If the player knows that reward are bounded in an interval $[a,b]$, and if she knows $a$ and $b$ ($a<b$), then with no loss of generality we can restrict to the interval $[0,1]$, as if $r\in[a,b]$, the player can instead consider the normalized reward $r' = $ that lies in $[0,1]$.
Note that this ``normalization trick'' is implemented for any policy in SMPyBandits, with the \texttt{lower} and \texttt{amplitude} optional arguments, respectively representing $a$ the lower-bound on rewards and $b-a$ the length of the interval of possible values of rewards.

@@

\paragraph{On one-dimensional exponential family.}
%
On one-dimensional exponential family...

@@

\TODOL{On garde ce petit paragraphe, par curiosité ?}

@@

@@

    \item
    - hyper-parameters tuning of deep neural networks etc
\end{itemize}

@@

\paragraph{Application for Opportunistic Spectrum Access.}
% Detail the previous work from our team SCEE on bandits + OSA : Wassim, Navik
%
Previous works of our SCEE team showed that MAB can be used to model the problem of spectrum access for a secondary user accessing a licensed spectrum.
In this model, arms represent a finite set of orthogonal channels, \ie, different frequency bands in a licensed spectrum.
In the model with sensing, the rewards $r_k(t)$ represents the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$.
A reward of $r_k(t) = 1$ indicates that no Primary User was sensed, while a reward of $r_k(t)=0$ indicates that the channel $k$ is busy at time $t$ and no uplink message can be sent.
%
This model was first introduced by Wassim Jouini during his PhD thesis,
in \cite{Jouini09} and later studied in both \cite{Jouini10,Jouini12}.
Proof-of-concepts using real-world radio hardware was first propsed in \cite{MoyWSR2014,RobertSDR2014}.
In a second PhD thesis supervised by Christophe Moy, Navikkumar Modi studied the impact
of using MAB algorithms to optimize channel selection
on the battery life of a wireless device.
On the one hand, running a MAB algorithm such as UCB-like algorithms was proved to be useful and can bring significant improvement in terms of successful transmission rates, directly increasing the battery life of the device.
On the other hand, classical MAB algorithms tend to switch arms a lot of times, especially in the beginning of the learning process, and this induces a lot of dynamical hardware reconfiguration for the wireless device, as selecting a different channel requires a change in the radio hardware used by the device.
Each hardware reconfiguration costs energy for the device, and quickly switching algorithms will lead to a reduction of the battery life.
The tradeoff between the two aspects is studied empirically in
\cite{modiDemo2016}.
%
Another interesting works is \cite{Modi17QoS}, FIXME

>>>>

\paragraph{Abstract.}
%
In this chapter, we present the common base of all the models studied in this thesis:
the multi-armed bandit (MAB) model,
restricting to the single-player and stationary stochastic case.
We focus on decision-making models with a finite number of resources, called arms, and on stochastic models, where an arm is associated with a one-dimensional distribution.
In all this thesis, we only consider one-dimensional exponential families, and in particular Bernoulli distributions.
%
We review quickly the possible applications of bandits, that range from recommender systems and information retrieval to healthcare and finance. In the rest of this thesis, we focus on cognitive radio, an important field of research where bandits has already been applied with success.

We then explain the measure of performance used in this thesis, defined as the regret and written $R_T^{\cA}$ for algorithm $\cA$ and horizon $T$. The regret compares the performance of the decision-making player and the oracle who knows in advance the best arm and always selects it.
Other measures of performance, such as best-arm identification, are not used later on in this document, but they are also shortly discussed.
%
We review the main theoretical results from famous previous works, that give a lower-bound on the regret of any algorithm, essentially showing that we cannot find an algorithm achieving a sub-logarithmic regret, in this stochastic MAB model ($R_T^{\cA} = \Omega(\log(T))$).

We give an overview of the main families of algorithms designed to tackle the stochastic MAB model, and we mainly focus on the \UCB, \klUCB{} and Thompson sampling algorithms, as they are used in the rest of the manuscript.
When presenting our library SMPyBandits in next Chapter~\ref{chapter:3}, designed to perform numerical simulations of MAB problem, we use it to compare the empirical performances of the most famous and most efficient algorithms, in terms of regret and by evaluating them on different problems.
%
In the point-of-view of numerical simulations or real-world applications of bandits on embedded systems, one must also take care of two other measures of performance: time and memory complexities, which can vary greatly between different families of algorithms.
We also give an empirical comparison of different algorithms for these two measures in Section~\ref{sec:3:timeAndMemoryCosts}.

We also give regret upper-bounds of the two main algorithms used in this thesis, \UCB{} and \klUCB, to highlight the difference between their finite-time regret upper-bounds. They both attain a logarithmic regret, proving their order optimality (\ie, their regret upper-bound asymptotically matches the lower-bound up-to a constant factor). Moreover, \klUCB{} is known to be optimal for a wide range of arms distributions.
% A sketch of the proof of both results is given in Appendix.

@@

\paragraph{Main references for the curious reader.}
%
For more details and a more formal introduction to multi-armed bandits, we suggest to the interested reader to work on a very recent text-book by Slivkins \cite{Slivkins2019}.
Another excellent but reasonably short survey is the book by Bubeck and Cesa-Bianchi \cite{Bubeck12}, while the more recent book by Lattimore and Szepesv{\'a}ri \cite{LattimoreBanditAlgorithmsBook} is the most complete resource about bandit algorithms.
Finally, we recommend \cite{bouneffouf2019survey} for a short but good survey on applications of MAB.

@@

Multi-Armed Bandits models were first introduced by Thompson as early as 1933 in \cite{Thompson33}, and later studied from the 1950s by Robbins \cite{Robbins52} and others.
This family of models was first proposed for clinical trials, and later applied to a wide range of different problems.
Their name comes from the one-armed bandit found in casinos, see Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
% The seminal paper by Lai and Robbins \cite{LaiRobbins85} was the first to introduce the measure of performance widely used in the literature, the notion of \emph{regret}, and to analyze a lower-bound on the performance of any algorithm against a fixed problem.
% Since then, many efficient algorithms were proposed, XXX

A MAB is a model of a decision-making game, where a \emph{player} has to sequentially select one action in a (usually finite) set of actions, and only receives a (random) feedback about the selected action, also called a \emph{reward}.
We consider $K\in\N, K\geq2$ arms, and discrete times $t\in\N^*$.
When the player decides at time $t$ to play (or pull) the arm $A(t)\in\{1,\dots,K\}=[K]$, she receives a reward $r_{A(t)}(t)\in\R$, and so on for a finite number of steps $t$, from $t=1$ until $t=T$ for an horizon $T$.
%
A commonly studied goal for the player is to maximize its sum of received rewards, $\sum_{t=1}^T r_{A(t)}(t)$
(or its expectancy in stochastic and stationary models).

From a mathematical point-of-view, this quantity $\sum_{t=1}^T r_{A(t)}(t)$ is indeed random, and it depends on two aspects.
First, the cumulated rewards depend on the randomness from the environment, that is, the unknown and unpredictable values of $r_k(t)$ for all $k$ and $t$.
Then, this quantity also depends on the decision-making policy followed the player, that is, the choices $A(t)$, that are based on (all) the past observations ($A(1), r_{A(1)}(1), \dots, A(t-1), r_{A(t-1)}(t-1)$), and possibly on an external source of randomness.
Without loss of generality, it can be given by a sequence $U_0,U_1,\dots$ of \iid{} random values, following the uniform distribution on $[0,1]$, and that has to be independent from the rewards $r_k(t)$.
This source of randomness is used in most MAB algorithm, for instance to break ties in a uniformly random manner.
%  in the decision making rule.

\tikzstyle{block} = [align=center, draw, fill=gray!25, rectangle, minimum height=3em, minimum width=6em]
\begin{figure}[h!]
    \centering
    \resizebox{0.50\textwidth}{!}{
        \begin{tikzpicture}[auto, node distance=5cm, >=latex]
            %
            % We start by placing the blocks
            \node [block] (player) at (0,0) {Player};
            % We draw an edge between the player and system block to
            \node [block] (environment) at (4,0) {MAB problem};
            % Once the nodes are placed, connecting them is easy.
            \draw [->] (player) to[bend left=90] node[pos=0.5] {Chooses a discrete action $k=A(t) \in\{1,\dots,K\}$} (environment);
            \draw [->] (environment) to[bend left=90] node[pos=0.5] {Observes a real-valued reward $r_k(t) \in \R$} (player);
            %
        \end{tikzpicture}
    }
\caption{Reinforcement learning cycle in a MAB model, for time steps $t=1,\dots,T$.}
\label{fig:2:ReinforcementLearningCycleMABmodel}
\end{figure}

The difficulty of this decision-making game comes from balancing the trade-off between \emph{exploration}, because the player has to observe as much as possible all the arms to get more knowledge about the distributions of their rewards, and \emph{exploitation}, because she also has to select as much as possible the best arm.
The MAB model is a famous example of a reinforcement learning model, where the decision-making process has to adapt to an unknown environment using noisy observations and a discrete time, alternating between decisions and observations.
We illustrate this cycle in Figure~\ref{fig:2:ReinforcementLearningCycleMABmodel} below.
For more details on reinforcement learning, in particular about more generic models, we refer to the famous book \cite{SuttonBarto2018}.

@@

In all this thesis, we focus on the model with finite arms, and binary or real-valued stochastic rewards, that is $A(t)\in\{1,\dots,K\}$ for a fixed and known number of arms $K\in\N,K\geq2$, and $r_{A(t)}(t)\in\R$.
%
Restricting to models with a finite number of arms
rules out an entire domain of the recent research on MAB, in particular we do not consider linear nor contextual bandits (for more details, see Parts V and VI of \cite{LattimoreBanditAlgorithmsBook}).
%
We also restrict to \emph{stochastic} rewards, that is we associate a real-valued distribution to each arm, denoted $\nu_k$ for arm $k\in\{1,\dots,\}=[K]$.
Rewards are stationary, meaning that $(r_k(t))_{t\in\N^*}$ is independent and identically distributed (\iid), $r_k(t) \sim \nu_k$ for any $t\geq1$.

@@

@@

\begin{figure}[h!]  % [htbp]
    \centering
    \includegraphics[width=0.85\linewidth]{2-Chapters/2-Chapter/Images/example_of_a_5_arm_bandit_problem.png}
    \caption{Screenshot of the demonstration available online on my website, for a current step of $t=24$.}
    \label{fig:2:example_of_a_5_arm_bandit_problem}
\end{figure}

The webpage looks like Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} above.
The arms follow Bernoulli distributions, \ie, $\nu_k = \mathcal{B}(\mu_k)$, of unknown means $\mu_k\in[0,1]$, and your goal in this interactive demonstration is to obtain the highest possible cumulated reward in $100$ steps, $\sum_{t=1}^{100} r_{A(t)(t)$.
Your decisions have to be made sequentially: at time $t$, you pick one of the arms, $A(t) \in\{1,2,3,4,5\}$, then the demo shows the random reward obtained from this (virtual) casino machine (in \textcolor{gold}{yellow}), \ie, the binary value $r_{A(t)}(t)\in\{0,1}$, that is sampled \iid{} from $\nu_k$.
%
The UI of the demo also shows the current value of $t$ (``total plays'') and $\sum_{s=1}^t r_{A(s)}(s)$ the ``total reward''.
For each arm, we show the sum of rewards obtained from that arm, \ie, $X_k(t) = \sum_{s=1}^t r_k(s) \mathbbm{1}(A(s) = k)$, in the ``Rewards'' line, and the number of pulls of that arm, \ie, $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$ in the ``Pulls'' line.
%
The demo also shows the estimated probability of each arm, that is $\widehat{\mu_k}(t) = X_k(t) / N_k(t)$ (when $N_k(t)>0$), in the ``Estimated probs'' line.

In the first Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem}, the current state of the game is shown at time $t=24$.
% Currently at time $t=24$ (out of $T=100$ total time steps),
At this step, the player has collected a sum of rewards of $14$, by observing $X(t) = [6,2,2,2,2]$ rewards of value $1$ in the $K=5$ different arms. Arms were sampled $N(t) = [8,4,4,4,4]$ times, meaning that the value $0$ was seen respectively $[2,2,2,2,2]$ times, and currently arm $1$ appears to be the best one. The true means of the arms are $\bm{\mu}=[0.6, 0.2, 0.55, 0.7, 0.5]$, and (much) more samples are needed before the player can accurately identify arm $4$ as the best arm.
%
In the second Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem__step100} below, we display the result of an example of run, when the player was following the \UCB{} algorithm from \cite{Auer02}.
After $T=100$ steps, the player obtained a cumulated reward of $56$, by playing mostly arms $4,3,5,1,0$ (in decreasing order of number of plays). The empirical means $\widehat{\mu_k}(T)$ correctly identify the best arm (arm $4$), but do not correctly rank the arms as arms $1$ and $3$ obtained means of $\widehat{\mu_1}(T) = 0.5 < \widehat{\mu_3}(T)=0.6$ but the true means are $\mu_3 = 0.55 < \mu_1 = 0.6$.

\begin{figure}[h!]  % [htbp]
    \centering
    \includegraphics[width=0.95\linewidth]{2-Chapters/2-Chapter/Images/example_of_a_5_arm_bandit_problem__step100.png}
    \caption{Screenshot of the demonstration, after the end of the game, after $t=T=100$ steps.  See \href{https://perso.crans.org/besson/phd/MAB\_interactive\_demo/}{\texttt{perso.crans.org/besson/phd/MAB\_interactive\_demo/}}}
    \label{fig:2:example_of_a_5_arm_bandit_problem__step100}
\end{figure}

\paragraph{A bandit policy is a measurable function.}

A bandit algorithm $\cA$, also referred to as a strategy or a policy, and sometimes denoted by $\pi$ or $\rho$, selects an arm $A(t)$ at time $t$ by using the past observations and the past external randomness.
More formally, an algorithm is thus a sequence of measurable functions $(\cA_t)_{t\geq1}$,
where $\cA_t$ maps the past observations $\cO_t = (U_0, Y_1, U_1, \dots, Y_{t-1}, U_{t-1})$
to an arm $\cA_t(\cO_t) = A(t) \in[K]$,
if we denote $Y_s = r_{A(s)}(s)$ she $s$-th reward.
The initial information is reduced to $\cO_1 = (U_0)$, and the first decision is $A(1) = \cA_1(\cO_1)$. Usually, most algorithms starts by selecting $A(1)=1,\dots,A(K)=K$ (or a permutation of the $K$ arms) in the $K$ first steps.
%
An algorithm is said to be deterministic if it does not depend on the external randomness $U_0,U_1,\dots$, but in this thesis we only use non-deterministic strategies.

The most common objective for the player is to maximize its expected rewards, that is $\E[\sum_{t=1}^T Y_t] = \E[\sum_{t=1}^T r_{A(t)}(t)]$.
The expectancy is taken on the randomness of the rewards $(r_k(t))_{k,t}$ as well as the randomness $(U_s)_s$.

\TODOL{A finir, a faire comme dans https://tel.archives-ouvertes.fr/tel-02121614/document page 19}

@@

% Just bounded, in $[a,b]$ but without loss of generality in $[0,1]$ etc.
\paragraph{The normalization trick.}
%
If the player knows that reward are bounded in an interval $[a,b]$, and if she knows $a$ and $b$ ($a<b$), then with no loss of generality we can restrict to the interval $[0,1]$, as if $r\in[a,b]$, the player can instead consider the normalized reward $r' = $ that lies in $[0,1]$.
Note that this ``normalization trick'' is implemented for any policy in SMPyBandits, with the \texttt{lower} and \texttt{amplitude} optional arguments, respectively representing $a$ the lower-bound on rewards and $b-a$ the length of the interval of possible values of rewards.

@@

\paragraph{On one-dimensional exponential family.}
%
% On one-dimensional exponential family...

In all this thesis we restrict to distributions that have a certain form, and focus on real-valued distributions lying in an exponential families.
One-dimensional exponential families include Bernoulli and Poisson distributions, as well as Gaussian distributions with a fixed variance.
Our main interest is Bernoulli distributions, but we prefer to present here the notations on a generic exponential family.
We follow the notations from the course on statistical learning (Stat 260) taught in 2010 by Michael Jordan at the University of Berkeley \cite{JordanCourseStatBerkeley}, that are also the notations used for instance in Emilie Kaufmann's PhD in 2014 \cite{Kaufmann12PhD}.

Given a measure $\lambda$, that is usually the natural Lebesgue measure on $\R$, an exponential family of probability distributions is defined as the distributions whose density (relative to $\lambda$) have the following form:
\[ \Pr(x | \lambda) = h(x) \exp(\eta^T T(x) - A(\eta)), \]
for a parameter vector $\eta$ (the canonical parameter), and given functions $T$ (the sufficient statistic) and $h$.
$A(\eta)$ is the cumulant function, and it is entirely determined by the $\eta$, $h$, $T$,
$A(\eta) = \log \int h(x) \exp(\eta^T T(x)) \lambda(\mathrm{d} x)$.
%
The natural parameter space is the set of values of $\eta$ such that this integral $A(\eta)$ is finite,
and usually the literature focusses on regular and minimal exponential families (when the natural parameter space is an non-empty open set in $\R$).

An important result on exponential families is that a distribution in such family is entirely characterized by its parameter $\eta$.
That is why such one-dimensional distributions are entirely characterized by their mean, as $\mu = \E_{\eta}[T(X)]$.

A second important result is a simplified form for the Kullback-Leibler divergence \cite{KullbackLeibler51}, for two distributions lying in the same exponential family.
The KL divergence between two distributions $d_1=\Pr(x|\eta_1)$ and $d_2=\Pr(x|\eta_2)$ is defined as
\[ \KL(\Pr(x|\eta_1), \Pr(x|\eta_2)) = \int \Pr(x|\eta_1) \log(\frac{\Pr(x|\eta_1)}{\Pr(x|\eta_1)}) \lambda(\mathrm{d}x) = \E_{d_1}[ \log(\frac{\Pr(x|\eta_1)}{\Pr(x|\eta_1)}) ]. \]
%
So it can be written $\KL( \eta_1,  \eta_2)) = \KL(\Pr(x|\eta_1), \Pr(x|\eta_2))$ and simplified to
\[ \KL(\Pr(x|\eta_1), \Pr(x|\eta_2)) = \E_{d_1}[ (\eta_1 - \eta_2)^T T(X) ] - A(\eta_1) + A(\eta_2) = (\eta_1 - \eta_2)^T \mu_1 - A(\eta_1) + A(\eta_2), \]
where $\mu_1 = \E_{\eta_1}[T(X)]$ is the mean of distribution $d_1$.

Without diving too much in the details, it is interesting to illustrate these notations for two examples.
On the first hand, Bernoulli distributions can be seen as an exponential family with
$T(x) = x$ and $h(x) = 1$,
and for a Bernoulli distribution of mean $\mu\in[0,1]$,
the parameter is $\eta = \mu / (1 - \mu)$, giving $A(\eta) = \log(1 + \mathrm{e}^{\eta})$
(with the limit behavior $\eta=+\infty$ if $\mu=0$).
%
On the other hand, Gaussian distributions use
$T(x) = [x ; x^2]^T$ and $h(x) = 1/\sqrt(2\pi)$,
and for $\cN(\mu,\sigma^2)$ a Gaussian distribution with mean $\mu\in\R$ and variance $\sigma^2$,
$\eta = [\mu/\sigma^2 ; -1/\sigma^2]^T$, giving $A(\eta) = -\eta_1^2/4\eta_2 -1/2 \log(-2\eta_2)$ (see Chapter~8 of \cite{JordanCourseStatBerkeley}).
Thus fixing the variance $\sigma$ gives indeed a one-dimensional exponential family.

In the rest of this thesis, we only consider bounded rewards, and we mostly focus on Bernoulli distributions.
First, we restrict for simplicity to Bernoulli distributions for the simulations shown in Chapter~\ref{chapter:3}, even if our library does implement all the distributions commonly found in the literature (including unbounded one-dimensional exponential families and even Markov models).
% Then in Chapter~\ref{chapter:4}, each static IoT device in the noisy environment is assumed to be following a Bernoulli emission pattern in a fixed channel, FIXME
%
For example when we study multi-player bandits in Chapter~\ref{chapter:5} (for the model without sensing) we explain that restricting to the Bernoulli case is interesting as it the harder one, as continuous distributions with a null mass on $r(t)=0$ yield a much simpler problem.
%
Finally, in Chapter~\ref{chapter:6} we analyze our proposed algorithm in the setting on bounded distributions, without restricting to Bernoulli distributions, but we use the fact that bounded distributions on $[0,1]$ are sub-Bernoulli, and we use Bernoulli Kullback-Leibler divergences in our analysis.

@@

\TODOL{On garde ce petit paragraphe sur les modèles de bandits Markovien, par curiosité ?}

@@

The first application of MAB models was for clinical trials, were an arm represent a treatment, and the distribution associated with such treatment can be a Bernoulli distribution: a reward of $0$ means the drug did not heal the disease, and a reward of $1$ indicates a success. The mean of an arm, in this application, represents the mean success rate of a treatment, and the goal of a doctor in a clinical trial is to identify the best treatment, \ie, the arm with highest mean, in a number of trials as short as possible.
%
The focus of this work is on cognitive radio and IoT networks, where arms can represent wireless orthogonal channels, but more generally any resource characterizing the communication of a device to a gateway (\eg, spreading factor for LoRa, power allocation for NOMA etc).
%
% After introducing formally the notations used in all this thesis, we review below possible applications of multi-armed bandits.

@@

    \item
    Using bandit algorithms to improve machine learning models or algorithms has been an active research domain for the last ten years or so.
    As presented below in Section~\ref{sec:2:chooseYourPreferredBanditAlgorithm}, a certain ``master'' bandit algorithm can be used to select on the run the best bandit algorithm from a pool of ``slave'' algorithms.
    Other possible use cases include hyper-parameter optimization, or feature selection.
    Hyper-parameters include real-valued parameters, like $\alpha\in[0,\infty)$ for the \UCB{} or $\varepsilon$ for the $\varepsilon$-greedy bandit algorithms, a step size multiplier in a gradient method, or the width of a radial based function (RBF) kernel method.
    Discrete-valued parameters are also common, like a choice in a fixed set of kernel functions or the depth of neural networks,
    and higher dimensional or more complex hyper-parameters include the entire architecture of a neural network.

    % FIXME this is copied pasted from \cite{bouneffouf2019survey}
    It is well known that performance of machine learning algorithms depends critically on identifying a good set of hyper-parameters.
    While recent approaches use Bayesian optimizationto adaptively select optimal hyper-parameter configurations, they rather focus on speeding up random search
    through adaptive resource allocation and early-stopping.
    A different approach was initiated in \cite{LiJamieson2018}, by formulating hyper-parameter optimization as a pure-exploration non-stochastic infinite-armed bandit problem, where a predefined resources, such as iterations, data samples, or features are allocated to randomly sampled configurations. This work introduced a novel algorithm, Hyperband, for this framework and analyze its theoretical properties, providing several desirable guarantees.
    Furthermore, Hyperband was compared with popular Bayesian optimization methods on a suite of hyper-parameter optimization problems; it was observed that Hyperband can provide more than an order-of-magnitude speedup over its competitors on a variety of deep-learning and kernel-based learning problems.

\end{itemize}

@@

\paragraph{Application for Opportunistic Spectrum Access.}
% Detail the previous work from our team SCEE on bandits + OSA : Wassim, Navik
%
Previous works of our SCEE team showed that MAB can be used to model the problem of spectrum access for a secondary user accessing a licensed spectrum.
In this model, arms represent a finite set of orthogonal channels, \ie, different frequency bands in a licensed spectrum.
In the model with sensing, the rewards $r_k(t)$ represents the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$.
A reward of $r_k(t) = 1$ indicates that no Primary User was sensed, while a reward of $r_k(t)=0$ indicates that the channel $k$ is busy at time $t$ and no uplink message can be sent.
%
This model was first introduced by Wassim Jouini during his PhD thesis,
in \cite{Jouini09} and later studied in both \cite{Jouini10,Jouini12}.
Proof-of-concepts using real-world radio hardware was first propsed in \cite{MoyWSR2014,RobertSDR2014}.
In a second PhD thesis supervised by Christophe Moy, Navikkumar Modi studied the impact
of using MAB algorithms to optimize channel selection
on the battery life of a wireless device.
On the one hand, running a MAB algorithm such as UCB-like algorithms was proved to be useful and can bring significant improvement in terms of successful transmission rates, directly increasing the battery life of the device.
On the other hand, classical MAB algorithms tend to switch arms a lot of times, especially in the beginning of the learning process, and this induces a lot of dynamical hardware reconfiguration for the wireless device, as selecting a different channel requires a change in the radio hardware used by the device.
Each hardware reconfiguration costs energy for the device, and quickly switching algorithms will lead to a reduction of the battery life.
The tradeoff between the two aspects is studied empirically in
\cite{modiDemo2016}.
%
Another interesting works is \cite{Modi17QoS}.
\TODOL{FIXME Finir ce blabla inutile ? Ou alors ce sera déjà fait en intro/chapitre1 ??}
